Exponential methods for solving hyperbolic problems with application to kinetic equations

### 2020-12-04 communication/20-12-04_canum/canum.tex
🖥  final version of canum slides

@@

  Linear part is exact! \cmark
\end{frame}
%-------%
\begin{frame}{Idea of exponential integrators}
  \textbf{\color{mblue} 2 classes of methods:}
  \begin{description}[leftmargin=0.25cm]
    \item[\mbold{exponential Runge-Kutta:}] solve exactly what we can, and interpolate the rest. For example first order exponential Euler method:
      $$
        u(t_n+\Delta t) \approx u^{n+1} = e^{-ia\Delta t}u^n + \Delta t\varphi_1(ia\Delta t)F(u^n)
      $$
      where $\varphi_1(z) = \dfrac{e^z - 1}{z}$
      \vspace{-0.1cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Hochbruck:2010}
      \end{thebibliography}
    \item[\mbold{Lawson:}] Change of variable: $v(t)=e^{-iat}u(t)$, we solve with a RK method: $\dot{v} = \tilde{F}(t,v) = e^{-iat}F(e^{iat}v(t))$

      For example, Lawson Euler method:
      $$
        v(t_n+\Delta t)\approx v^{n+1} = v^n + \Delta t e^{-iat_n}F(e^{iat_n}v^n)
      $$
      or as an expression of $u$:
      $$
        u^{n+1} = e^{-ia\Delta t}u^n + \Delta te^{ia\Delta t}F(u^n)
      $$
      \vspace{-0.75cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Isherwood:2018}
      \end{thebibliography}
  \end{description}
\end{frame}

\section{Linear analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Reminder of stability tools}
  \only<1>{
    If we want to study stability of:
    $$
      \partial_t u + \partial_x u = 0
    $$
    with centered scheme (CD2) $(\partial_xu)_j \approx \frac{1}{2\Delta x}(u_{j+1}-u_{j-1})$. After a Fourier transform (\emph{von Neumann analysis}):
    $$
      \dot{u} + i\frac{\sin(k\Delta x)}{\Delta x}u = 0
    $$
    Explicit Euler method in time: we have to stretch \mbold{eigenvalues} (or \mbold{Fourier symbol}) of CD2 into explicit Euler \mbold{stability domain}.
  } \only<2>{
    \begin{figure}\centering
      \includegraphics[height=0.8\textheight]{img/cfl_example.png}
    \end{figure}
  }
\end{frame}
%-------%
\begin{frame}{From linear Vlasov equation to toy model}
  Linear Vlasov equation:
  $$
    \partial_t f + a\partial_x f + b\partial_v f = 0
  $$
  Fourier transform in $x$, CD2 in $v$ plus a Fourier transform in $v$, formally:
  $$
    \frac{\mathrm{d}f}{\mathrm{d}t} + iak f + b\frac{i\sin(\varphi)}{\Delta x}f = 0
  $$
  \mbold{Toy model:}
  $$
    \dot{u} + iau + \lambda u = 0
  $$
  with $a\in\mathbb{R}$, $\lambda\in\mathbb{C}$ (diffusive scheme for example).

@@

\subsection{Lawson methods}
%------------------------------------------------------------------------------
\begin{frame}{Lawson methods stability domain}
  For our toy model:
  $$
    \dot{u} = iau + \lambda u
  $$
  Change of variable: $v(t) = e^{-iat}u(t)$
  $$
    \dot{v} = e^{-iat}\lambda e^{iat}v
  $$
  Apply a Runge-Kutta method to compute stability function of Lawson method:
  $$
    v^{n+1} = \underbrace{p(\lambda\Delta t)}_{\text{stability function of RK}}v^n
  $$
  \emph{i.e.}:
  $$
    u^{n+1} = \overbrace{p(\lambda\Delta t)e^{-ia\Delta t}}^{\text{stability function of Lawson}}u^n
  $$
  Stability domain: $\mathcal{D}=\left\{z\in\mathbb{C},|p(z)|\leq 1\right\}$ of Lawson method is \textbf{the same} as the underlying Runge-Kutta method \textbf{because} $ia\in i\mathbb{R}$
\end{frame}
%-------%
\begin{frame}{Considered $Lawson(RK(s,p))$ methods}
  \begin{figure}\centering
    \includegraphics[height=0.75\textheight]{img/rk_sd.png}
    \caption{$\{z\in\mathbb{C},|p(z)|=1\}$}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2}
  For stability between a Lawson method and CD2, we solve:
  $$
    |p(iy)| = 1,\quad y\in\mathbb{R}
  $$
  \begin{figure}\centering
    \includegraphics[width=0.8\textwidth]{img/yaxis.png}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- LW5}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_scheme}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_rk44_weno.png}
    \end{column}
  \end{columns}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2/LW5: CFL estimates}
  \begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline
      - & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
      \hline
      CD2 ($y_{\max}$) & $\sqrt{3}$ & $2\sqrt{2}$\\
      \hline
      LW5 ($\sigma$) & $1.433$   & $1.73$   \\
      \hline  
    \end{tabular}
    \caption{CFL number for some Lawson schemes.}
  \end{table}
  Other RK--CD2 CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Baldauf:2008}
\end{thebibliography}
  Same results for RK--WENO CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Motamed:2010}
  \bibitem{} \customcite{Lunet:2017}
\end{thebibliography}
\end{frame}

>>>>

\usepackage{hyperref}

@@

  Linear part is exact! \cmark
\end{frame}
%-------%
\begin{frame}{Idea of exponential integrators}
  \textbf{\color{mblue} 2 classes of methods:}
  \begin{description}[leftmargin=0.25cm]
    \item[\mbold{exponential Runge-Kutta:}] solve exactly what we can, and interpolate the rest. For example first order exponential Euler method:
      $$
        u(t_n+\Delta t) \approx u^{n+1} = e^{-ia\Delta t}u^n + \Delta t\varphi_1(ia\Delta t)F(u^n)
      $$
      where $\varphi_1(z) = \frac{1}{z}\int_0^z e^{z-\tau}\,\mathrm{d}\tau = \dfrac{e^z - 1}{z}$
      \vspace{-0.1cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Hochbruck:2010}
      \end{thebibliography}
    \item[\mbold{Lawson:}] Change of variable: $\tilde{u}(t)=e^{-iat}u(t)$, we solve with a RK method: $\dot{\tilde{u}} = \tilde{F}(t,\tilde{u}) = e^{-iat}F(e^{iat}\tilde{u}(t))$

      For example, Lawson Euler method:
      $$
        \tilde{u}(t_n+\Delta t)\approx \tilde{u}^{n+1} = \tilde{u}^n + \Delta t e^{-iat_n}F(e^{iat_n}\tilde{u}^n)
      $$
      or as an expression of $u$:
      $$
        u^{n+1} = e^{-ia\Delta t}u^n + \Delta te^{ia\Delta t}F(u^n)
      $$
      \vspace{-0.75cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Isherwood:2018}
      \end{thebibliography}
  \end{description}
\end{frame}

\section{Linear analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Reminder of stability tools}
  \only<1>{
    If we want to study stability of:
    $$
      \partial_t u + \partial_x u = 0
    $$
    with centered scheme (CD2) $(\partial_xu)_j \approx \frac{1}{2\Delta x}(u_{j+1}-u_{j-1})$. After a Fourier transform (\emph{von Neumann analysis}):
    $$
      \dot{u} + i\frac{\sin(k\Delta x)}{\Delta x}u = 0
    $$
    Explicit Euler method in time: we have to stretch \mbold{eigenvalues} (or \mbold{Fourier symbol}) of CD2 into explicit Euler \mbold{stability domain}.
  } \only<2>{
    \begin{figure}\centering
      \includegraphics[height=0.7\textheight]{img/cfl_example.png}
    \end{figure}
    \begin{itemize}
      \item Stability function $p(z)$: $u^{n+1} = p(\Delta t\lambda)u^n$ to solve $\dot{u} = \lambda u$
      \item Stability domain: $\mathcal{D}=\left\{z\in\mathbb{C},|p(z)|\leq 1\right\}$
    \end{itemize} 
  }
\end{frame}
%-------%
\begin{frame}{From linear Vlasov equation to toy model}
  Linear Vlasov equation:
  $$
    \partial_t f + a\partial_x f + b\partial_v f = 0
  $$
  Fourier transform in $x$, CD2 in $v$ plus a Fourier transform in $v$, formally:
  $$
    \frac{\mathrm{d}f}{\mathrm{d}t} + iak f + b\frac{i\sin(\varphi)}{\Delta x}f = 0
  $$
  \mbold{Toy model:}
  $$
    \dot{u} + iau + \lambda u = 0
  $$
  with $a\in\mathbb{R}$, $\lambda\in\mathbb{C}$ (diffusive scheme for example).

@@

\subsection{Lawson methods}
%------------------------------------------------------------------------------
\begin{frame}{Lawson methods stability domain}
  For our toy model:
  $$
    \dot{u} = iau + \lambda u
  $$
  Change of variable: $\tilde{u}(t) = e^{-iat}u(t)$
  $$
    \dot{\tilde{u}} = e^{-iat}\lambda e^{iat}\tilde{u}
  $$
  Apply a Runge-Kutta method to compute stability function of Lawson method:
  $$
    \tilde{u}^{n+1} = \underbrace{p(\lambda\Delta t)}_{\text{stability function of RK}}\tilde{u}^n
  $$
  \emph{i.e.}:
  $$
    u^{n+1} = \overbrace{p(\lambda\Delta t)e^{-ia\Delta t}}^{\text{stability function of Lawson}}u^n
  $$
  Stability domain: $\mathcal{D}=\left\{z\in\mathbb{C},|p(z)|\leq 1\right\}$ of Lawson method is \textbf{the same} as the underlying Runge-Kutta method \textbf{because} $ia\in i\mathbb{R}$
\end{frame}
%-------%
\begin{frame}{Considered $Lawson(RK(s,p))$ methods}
  \begin{figure}\centering
    \includegraphics[height=0.75\textheight]{img/rk_sd.png}
    \caption{$\{z\in\mathbb{C},|p(z)|=1\}$}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2}
  For stability between a Lawson method and CD2, we solve:
  $$
    |p(iy)| = 1,\quad y\in\mathbb{R}
  $$
  \begin{figure}\centering
    \includegraphics[width=0.8\textwidth]{img/yaxis.png}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- LW5}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_scheme}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_rk44_weno.png}
    \end{column}
  \end{columns}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2/LW5: CFL estimates}
  \begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline
      - & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
      \hline
      CD2 ($y_{\max}$) & $\sqrt{3}$ & $2\sqrt{2}$\\
      \hline
      LW5 ($\sigma$) & $1.433$   & $1.73$   \\
      \hline  
    \end{tabular}
    \caption{CFL number for some Lawson schemes.}
  \end{table}
  Other RK--CD2 CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Baldauf:2008}
\end{thebibliography}
  Same results for RK--WENO CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Motamed:2010}
  \bibitem{} \customcite{Lunet:2017}
\end{thebibliography}
  More results of CFL estimates:
  \url{http://jmassot.perso.math.cnrs.fr/ponio.html}
\end{frame}
